A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-22 jacobian_calc_CMAME.tex
added lit review to paper

\subsubsection{The ``complex-step'' method.}  \label{sec:CSmethod} It is
possible to approximate derivatives quite accurately with a technique based on
a complex variable Taylor series expansion of a function.  This method was
first described by Lyness and Moler
\cite{lyness1967numerical,lyness1968differentiation} and has more recently been
rediscovered
\cite{squire1998using,voorhees2011complex,al2010complex,martins2000automated}
for use in engineering analysis.  The basic idea is that a model parameter can
be be made complex and expanded in a Taylor series about a small perturbation
along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i
h}{1!}  \notag \\ &+ \frac{\partial f^2(x)}{\partial x^2} \frac{(i h)^2}{2!} +
\frac{\partial f^3(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots,
\label{eqn:complexTaylor} \end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor})
and solving for the first derivative and ignoring terms of $\mathcal{O}\left (
h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} \approx \frac{\mbox{Im}
\left( f (x + i h) \right)}{h}.  \label{eqn:complexFirstDeriv} \end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only
utilizing one functional evaluation of a perturbed model parameter along the
imaginary axis.  The step-size $h$ is arbitrary and can be made as small as
practical (even to machine precision without the dangers of roundoff error as
in finite-differences) to yield an accurate estimate of this derivative.  The
only disadvantage of this technique is the necessity of requiring the functions
to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a
simple example that illustrates the use of the complex-step method for the
derivative calculation of a function.

@@

\begin{figure}[tbp] \centering \includegraphics{./figs/truss.png}
\caption{Beam and spring simulated in example program.} \label{fig:TrussSchematic}
\end{figure}

>>>>

\subsubsection{The ``complex-step'' method.}  \label{sec:CSmethod} It is
possible to approximate derivatives quite accurately with a technique based on
a complex variable Taylor series expansion of a function.  This method was
first described by Lyness and Moler
\cite{lyness1967numerical,lyness1968differentiation} and has more recently been
rediscovered for use in engineering analysis.  The basic idea is that a model parameter can
be be made complex and expanded in a Taylor series about a small perturbation
along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i
h}{1!}  \notag \\ &+ \frac{\partial f^2(x)}{\partial x^2} \frac{(i h)^2}{2!} +
\frac{\partial f^3(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots,
\label{eqn:complexTaylor} \end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor})
and solving for the first derivative and ignoring terms of $\mathcal{O}\left (
h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} \approx \frac{\mbox{Im}
\left( f (x + i h) \right)}{h}.  \label{eqn:complexFirstDeriv} \end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only
utilizing one functional evaluation of a perturbed model parameter along the
imaginary axis.  The step-size $h$ is arbitrary and can be made as small as
practical (even to machine precision without the dangers of roundoff error as
in finite-differences) to yield an accurate estimate of this derivative.  The
only disadvantage of this technique is the necessity of requiring the functions
to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a
simple example that illustrates the use of the complex-step method for the
derivative calculation of a function.

\note[Michael]{re: I.4}
It is appropriate to detail the prior work discussing aspects of the
complex-step method in order to indicate what is new in the instant work.  The
CS method was introduced in a work by J. N. Lyness and C. B. Moler
\cite{lyness1967numerical}.  In this prior work, the CS method is identified in
an exact form and the comptuing of analytical derivatives is shown. This prior
work develops the CS method, while the instant work is about an application of
the CS method.

A follow-on work by J. N. Lyness demonstrates a 'truncated' approximate version
of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.
This form of the CS method is used in the instant work.
The prior work develops the CS method, while the instant work is about an application of the CS method.

A work by W. Squire and G. Trapp showns through empirical data that CS offers superior accuracy
to FD and CD \cite{squire1998using}. The instant work uses the CS method as shown in the
prior work and was motivated by the prior work's suggestion that CS is highly accurate. 
The instant work examines applications of the CS method beyond taking derivatives alone,
whereas the prior work does not.

A work by J. C. Newman, W. K. Anderson and L. D. Whitfield, discusses computing
sensitivities in a CFD analysis using complex step and compares the results to
central difference produced results \cite{newman1998}. The work is also notable
for identifying automatic differentiation as a competing method to CS and
suggesting that complex step could be used to compute Jacobians of residuals
for use in Newton-Kyrlov schemes for solving non-linear systems. The instant
work is novel with respect to the prior work because the instant work
investigates CS as applied to a numerical method for solving non-linear
systems, rather than as in the prior work investigating the derivatives
themselves produced with the CS method.

In a work by A. Perez-Foguet, A. Rodriguez-Ferran, A. Huerta, CS is used to
compute Jacobians for both integrating plastic stress to determine
elastoplastic modulii, but also for computing the Jacobian of the global
problem, both of which are an application of the CS method to the solution of
non-linear systems \cite{perez2000numerical}. These elastoplastic modulii are
used in computing the global Jacobian  and said to be 'algorithmically
consistent' in that they are consistent with the algorithm used to return the
predicted stress to the yield surface. Examples of interest include FEM
simulations on two plastic material models with complex hardening rules.
Comparing the prior work with the instant work, the prior work demonstrates CS
on a serial computer and discusses a classical solid mechanics formulation.
Automatic differentiation is not investigated, and CPU time measurements are not
taken. Scalability studies with a variety of equivalent, progressively more
refined meshes are not done. Scalability as a function of parallel computer
size is not investigated. The instant work is novel with respect to the prior work in
offering a discussion of CS as applied to peridynamics simulation on a parallel
computer, CPU time data, and comparison to automatic differentiation as a competing method.

Another work by A. Perez-Foguet, A. Rodriguez-Ferran, A. Huerta The prior work
applies CS in computing derivatives needed for integrating stress, obtaining
algorithmically consistent elastoplastic modulii, and obtaining the Jacobian of
the global residual, for FEM simulations using the the MRS-Lade material model
\cite{perez2012numerical}. In this prior work, CPU runtime is measured.
Comparing the prior work to the instant work, the prior work demonstrates CS on
a serial computer and discusses a classical solid mechanics formulation.
Automatic differentiation is not investigated.  Scalability studies with a
variety of equivalent, progressively more refined meshes are not done.
Scalability as a function of distributed system size is not investigated.  The
instant work is novel with respect to the prior work in offering a discussion
of CS as applied to peridynamics simulation on a parallel computer, and
comparison to automatic differentiation as a competing method.

A work by Joaquim R. R. A. Martins, Ilan M. Kroo and Juan J. Alonso
investigates using CS for computing aerodynamic sensitivities
\cite{martins2000automated}. Additionally higher order accuracy versions of CS
are discussed, as is the task of implementing complex-step in most widely used
general purpose programming languages. Also dicussed is the relative ease of
implementing a CS scheme versus an AD based scheme for computing sensitivities
in an established code. The instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the using the CS method to obtain sensitivities.

A work by W. K. Anderson, J. C. Newman, D. l. Whitfield and E. J. Nielsen discusses
the CS applied to computing sensitivities in CFD simulations \cite{anderson2001sensitivity}. CS is
compared to AD both in Fortran and C++ applications. In this prior work it is suggested that forward mode
AD is mathematically equivalent to first order CS. Forware mode AD isThe instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the derivatives themselves produced with the CS method.

A work by J. Martins, P. Sturdza, J. Alonso demonstrates that the CS method can
be performed by using complex datatypes for the variables of functions
comprising an analysis code. The analysis code can comprise a simulation and
derivatives of response variables called sensitivities can be computed. The
accuracy of this approach is compared to automatic differentiation
\cite{martins2003complex}. The instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the derivatives themselves produced with the CS method.

A work by  K. L. Lai and J. L. Crassidis discusses first and second order CS
methods for computing derivatives. Choosing optimal step size
\cite{lai2008extensions}. This prior work discusses computing first and second
order accurate derivatives with CS and choosing optimal step size.  The intant
work discusses using the derviatives computed with the CS method in a numerical
method for solving a non-linear system.

A work by A. H. Al-Mohy and N. J. Higham investigates the use of complex step
in computing Frechet derivatives of matix to matrix functions \cite{al2010complex}. To
compute the residual Jacobian in the peridynamics simulation discussed in the
instant work is analogous because it is computing a Frechet derivative of a
higher dimension function of higher dimension variables. Where the application
of the CS derivatives in the prior work is limited to a particular class of
abstract functions to determine the CS' suitability for use in condition number
estimation, the instant work applies CS to solving a mechanics problem. The
instant work is not about computing Frechet derivatives with CS but moreover using them
in a numerical method to solve a set of equations describing a physical system.

A work by W. Jin, B. H. Dennis, and B. P. Wang discusses using CS to compute
sensitivities in a solid mechanics FEM analysis \cite{jin2010improved}. In the
prior work the derivitives produced by CS themselves are the object of
interest, where the instant work uses CS in a numerical method for solving a
non-linear system.

A work by A. Voorhees, H. Millwater, and R. Bagley, where H. Millwater is an author of
the instant work, dicusses the background of the CS method and frames it as a subset
of Fourier Differentiation. CS and Fourier Differentiation are then applied to 
computing sensitivities within a solid mechanics FEA \cite{voorhees2011complex}. A notable
technique demonstrated by this prior work is the representation of complex valued
nodal components as additional degrees of freedom in the FE model, to be interpreted
by a matrix representation of multicomplex numbers during computations, yet without storing
the entire matrix. In the prior work, CS is discussed and related to a broader class of differentiation methods and used
to compute sensitivities. In the instant work, CS is used in a numerical method for solving non-linear systems.

In another work by H. Millwater, C. Quintana, J. Garza and David Wagner the
concept of incorporating complex values into the degrees of freedom of finite
elements is labelled ZFEM \cite{millwater2013application}. Notable in this work
is the discussion of creating user defined element types for use with the
popular \emph{ABAQUS} software suite which enable the use of ZFEM in that
software, \emph{ABAQUS} is a trademark of Dassault Systemes
\cite{systemes2012abaqus}. In this prior work the goal of using CS through the
ZFEM technique is to compute shape sensitivities, while in the instant work CS
is used in a numerical method for solving non-linear systems.

A work by R. Abreu, D. Stich, and J. Morales discusses computing derivatives
with a re-interpreted version of the CS method which features the combined use
of real and imaginary valued step components \cite{abreu2013generalization}.
This prior work suggests that the modified CS method can achieve extended
approximation accuracy up to fourth order.  The prior work discusses using the
CS method to compute derivatives and attempts to extend the method for that
application, where the derivatives themselves are the object of interest. In
contrast, the instant work discusses using the derivatives computed with CS in
a numerical method for solving a non-linear system.

The body of prior literature has introduced the application of CS to
single dimension and multidimensional derivatives, sensitivity analysis,
non-linear solution schemes, the extension of CS for higher accuracy, 
and implementation strategies for CS in new and existing codes.
With respect to the prior work dicussing the CS method, the primary novel
aspects of the instant work are then its dicussion of the application of CS to
peridynamics simulations run in \emph{Peridigm}, a parallel implementation of
CS and a compilation of CPU time data, where each of CS, AD, FD and CD are
compared.

@@

\begin{figure}[tbp] \centering \includegraphics[width=0.7\textwidth, height=30px]{./figs/truss.png}
\caption{Beam and spring simulated in example program.} \label{fig:TrussSchematic}
\end{figure}
